Embedding Intuitionistic into Classical Logic

### 2022-10-13 FoSSaCS/fossacs-EIICL.tex
typos and abstract

\documentclass[runningheads]{llncs}
\usepackage{amsmath, amsfonts}
\usepackage{wrapfig}
\usepackage{multicol}
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
	%
	\title{Embedding Intuitionistic into Classical Logic}
	%
	%\titlerunning{Abbreviated paper title}
	% If the paper title is too long for the running head, you can set
	% an abbreviated paper title here
	%
	\author{Alexander Pluska \and Florian Zuleger}
	%
	\authorrunning{A. Pluska and F. Zuleger}
	% First names are abbreviated in the running head.
	% If there are more than two authors, 'et al.' is used.
	%
	\institute{TU Wien, Vienna, Austria}
	%
	\maketitle
\begin{abstract}
The famous double negation translation~\cite{glivenko1929quelques,godel1933intuitionistischen} establishes an embedding from classical into intuitionistic logic. Curiously the reverse direction has not been covered in literature. Based on the work in~\cite{claessen2015sat} we establish a small model property for intuitionistic propositional logic which allows us to utilize its Kripke semantics to give a simple embedding into classical propositional logic and quantified boolean formulas. We then transfer this procedure to first-order logic which allows us to give a embedding of intuitionistic first-order logic into classical first-order-logic.
This allows a direct application of state-of-the-art classical first-order provers to the problem of determining intuitionistic validity.

@@

\begin{definition}
	A \emph{Kripke structure} $\mathcal K = (W, (v_w)_{w\in W})$ consists of a partially ordered set $W$ of \emph{worlds}, called the Kripke \emph{frame}, and a family of valuations $(v_w)_{w\in W}$ such that $v_u(A)\leq v_u(A)$ for all $u\leq w$ and propositional variables $A$ (this is called the \emph{persistency condition}).
	The model relation between the $v_u$ and formulas $\varphi$ is defined as before, except in the case of implications, where we set
	\begin{itemize}
		\item $v_u\models\varphi\to \psi$ iff for all $w\geq u$ we have $w\not\models\varphi$ or $u\models\psi$.
	\end{itemize}
	We say that $\varphi$ is satisfied at a world $u$ if $v_u\models\varphi$ and write $u\models\varphi$. If a formula is satisfied at every world then $\mathcal K$ is a \emph{model} for $\varphi$. A formula is valid if every Kripke structure is a model of it. We denote the set of valid formulas with \emph{IPC} (Intuitionistic Propositional Calculus).
\end{definition}
There are many classical tautologies which are not intuitionistically valid, e.g. the law of excluded middle $A\vee\neg A$. On the other hand any intuitionistic theorem is a classical one.

@@

\begin{definition}
	Let $\Sigma$ be a signature. A $\Sigma$-structure $\mathcal{M}$ consists of a non-empty set $M$, the \emph{domain} of $\mathcal{M}$, and an \emph{interpretation} $I$ that assigns
	\begin{itemize}
		\item to each $n$-ary function symbol $f\in\Sigma$ a $n$-ary function $f^I: M^n\to M$.
		\item to each $n$-ary relation symbol $R\in\Sigma$ a $n$-ary relation $R^I\subseteq M^n$.
	\end{itemize}
	A variable assignment $v$ is a function that assigns to each free variable an element $m\in M$. For each free variable $a$ and $m\in M$ we define $$v[m/a](b) = \begin{cases}
	m, &\text{if $b=a$,}\\
	v(b), &\text{otherwise.}
	\end{cases}$$
	Then terms are interpreted as follows:
	\begin{itemize}
		\item $a^{I, v} = v(a)$ for each free variable $a$.
		\item $f(t_1,\dots,t_n)^{I, v} = f^I(t_1^{I, v},\dots, t_n^{I, v})$ for each $n$-ary function symbol $f\in\Sigma$.
	\end{itemize}
	We define a model relation between pairs $\mathcal M, v$ and formulas $\varphi$ as follows
	\begin{itemize}
		\item $\mathcal M, v\not\models\bot$
		\item $\mathcal M, v\models R(t_1,\dots,t_n)$ iff $(t_1^{I, v},\dots,t_n^{I, v})\in R^I$, $\mathcal M, v\models s = t$ iff $s^{I, v} = t^{I, v}$.
		\item $\mathcal M, v\models \varphi\wedge \psi$ iff $\mathcal M, v\models\varphi$ and $\mathcal M, v\models\psi$
		\item $\mathcal M, v\models \varphi\vee\psi$ iff $\mathcal M, v\models\varphi$ or $\mathcal M, v\models\psi$.
		\item $\mathcal M, v\models \varphi\to\psi$ iff $\mathcal M, v\not\models\varphi$ or $\mathcal M, v\models\psi$.
		\item $\mathcal M, v\models\exists x\varphi$ iff there exists $m\in M$ such that $\mathcal M, v[m/a]\models\varphi[a/x]$ where a is a free variable that does not occur in $\varphi$.
		\item $\mathcal M, v\models\forall x\varphi$ iff for all $m\in M$ we have $\mathcal M, v[m/a]\models\varphi[a/x]$ where a is a free variable that does not occur in $\varphi$.
	\end{itemize}
	A $\Sigma$-structure $\mathcal M$ satisfies $\varphi$ if $\mathcal M, v\models\varphi$ for every $v$, we write $\mathcal M\models\varphi$. A formula is valid if every $\Sigma$-structure $\mathcal M$ satisfies it. We denote the set of valid formulas with \emph{CQC} (Classical Quantified Calculus).
\end{definition}

\begin{definition}
	A $\Sigma$-\emph{Kripke structure} $\mathcal{K}$ is a partially ordered set $W$, called the Kripke \emph{frame}, and a family of $\Sigma$-structures $(\mathcal{M}_w)_{w\in W}$ such that for $u\leq w$ we have $M_u\subseteq M_w$, $f^{I_w}|_{M_u} = f^{I_u}$ and $R^{I_w}|_{M_u} = R^{I_u}$.\footnote{Here $f|_M$ denotes the restriction of $f$ to $M$}
	The model relation between $\mathcal M_u$, $v$ (targeting $\mathcal M_u$) and formulas $\varphi$ is defined as before, except in the cases of implication and $\forall$-quantification, where we set
	\begin{itemize}
		\item $\mathcal M_u, v\models \varphi\to\psi$ iff for every $w\geq u$ we have $w, v\not\models\varphi$ or $w, v\models\psi$.
		\item $\mathcal M_u, v\models\forall x\varphi$ iff for every $w\geq u$ and $m\in M_w$ we have $w, v[m/a]\models\varphi[a/x]$ where a is a free variable that does not occur in $\varphi$.
	\end{itemize}
	We write $u, v\models\varphi$ if $\mathcal M_u, v\models \varphi$. $\mathcal{K}$ satisfies $\varphi$ if $u, v\models\varphi$ holds for every world $u$ and variable assignment $v$. $\varphi$ is valid if it is satisfied by every Kripke structure.
	$\varphi$ is valid for some frame (write $W\models\varphi$) if it is satisfied in every Krike structure with that frame. We denote the set of valid formulas with \emph{IQC} (Intuitionistic Quantified Calculus).
\end{definition}
In addition to propositional tautologies there are now sentences involving quantifiers which are classically valid but not intuitionistically, e.g., $\neg\forall x A(x)\to \exists x \neg A(x)$.

@@

\begin{definition}
	For formulas $\varphi$ we define the Skolemization $\varphi^S_Z$ and Herbrandization $\varphi^H_Z$ with respect to $Z$ by simultaneous induction as follows:
	\begin{itemize}
		\item $A^S_Z = A^H_Z = A$ for each atomic $A$.
		\item $(\varphi\circ\psi)^X_Z = \varphi^X_Z\circ\psi^X_Z$ for $\circ\in\{\wedge, \vee\}$, $X\in\{S, H\}$.
		\item $(\varphi\to\psi)^S_Z = \varphi^H_Z\to \psi^S_Z$\\$(\varphi\to\psi)^H_Z = \varphi^S_Z\to\psi^H_Z$.
		\item $(\forall x\varphi)^S_Z = \forall x(\varphi[a/x]^S_{Z\cup\{a\}}[x/a])$ where $a$ is a new free variable\\$(\forall x\varphi)^H_Z = \varphi[s(z_1,\dots,z_n)/x]^H_Z$ where $s$ is a new function, $\{z_1\dots z_n\} = Z$.
		\item $(\exists x\varphi)^S_Z = \varphi[s(z_1,\dots,z_n)/x]^S_Z$ where $s$ is a new function, $\{z_1\dots z_n\} = Z$\\$(\exists x\varphi)^H_Z = \exists x(\varphi[a/x]^H_{Z\cup\{a\}}[x/a])$ where $a$ is a new free variable.
	\end{itemize}
	Let $$\varphi^S = (\exists x_1\dots\exists x_n \varphi[x_1/a_1\dots x_n/a_n])^S_\emptyset$$ and $$\varphi^H = (\forall x_1\dots\forall x_n \varphi[x_1/a_1\dots x_n/a_n])^H_\emptyset$$ where $a_1,\dots,a_n$ are the free variables occurring in $\varphi$.
\end{definition}

@@

Since we could not find a proof of this result one is given in the appendix as Lemma~\ref{proof:first-order-normal-form}.

@@

We shall now give an overview of our main results and highlight the key arguments. We start with the propositional case. The key arguments are the same as in the first-order case, while it is technically simpler and less cluttered.

@@

The key property that allows us to reduce our encoding, which a priori targets first order logic, to propositional logic, is a reduction of counter-models, which is a proper dual to validity for intuitionistic logic. In particular, given $$\varphi = \left(\bigwedge\mathcal R\wedge\mathcal X\right)\to P$$ as in Lemma~\ref{lemma:propositional-normal-form} we establish:

@@

This allows us to embed intuitionistic propositional logic into classical propositional logic.

\begin{lemma}
	For each propositional formula $\varphi$ there is effectively a formula $\varphi^\#$ with $\mathcal |\varphi^\#|\in\mathcal O(|\mathcal R|\cdot\exp(|\mathcal X|\log(|\mathcal X|)))$ such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid.
\end{lemma}

The exact details can be found in Theorem~\ref{thm:prop-translation}. The essential mechanisms is the elimination $\forall$-quantifiers by enumeration of worlds in a sufficiently general Kripke Frame. Instead of expressing correctness at each world of the model we can express that the worlds along some path in the tree are correct. The branching of paths can be expressed through quantification in a quantified boolean formula (QBF) which gives us the following:

@@

Again, the details can be found in Section~\ref{sec:qbf}. Next we attempt to transport these results to the first order case. There are formulas for which only counter-examples with an infinite Kripke Frame exist. Therefore, there is no hope of completely eliminating terms representing worlds from the final embedding. This also means that our language necessarily becomes more complex. However a number of reductions is still possible.

>>>>

\documentclass[runningheads]{llncs}
\usepackage{amsmath, amsfonts}
\usepackage{wrapfig}
\usepackage{multicol}
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following two lines
% to display URLs in blue roman font according to Springer's eBook style:
%\usepackage{color}
%\renewcommand\UrlFont{\color{blue}\rmfamily}
%
\begin{document}
	%
	\title{Embedding Intuitionistic into Classical Logic}
	%
	%\titlerunning{Abbreviated paper title}
	% If the paper title is too long for the running head, you can set
	% an abbreviated paper title here
	%
	\author{Alexander Pluska \and Florian Zuleger}
	%
	\authorrunning{A. Pluska and F. Zuleger}
	% First names are abbreviated in the running head.
	% If there are more than two authors, 'et al.' is used.
	%
	\institute{TU Wien, Vienna, Austria}
	%
	\maketitle
\begin{abstract}
The famous double negation translation~\cite{glivenko1929quelques,godel1933intuitionistischen} establishes an embedding from classical into intuitionistic logic. Curiously the reverse direction has not been covered in literature. Based on the work in~\cite{claessen2015sat} we establish a small model property for intuitionistic propositional logic in order to utilize its Kripke semantics to give a simple embedding into classical propositional logic and quantified boolean formulas. We transfer there methods to the first order case and also give an embedding of intuitionistic first-order logic into classical first-order-logic.
This allows a direct application of state-of-the-art classical first-order provers to the problem of determining intuitionistic validity.

@@

\begin{definition}
	A \emph{Kripke structure} $\mathcal K = (W, (v_w)_{w\in W})$ consists of a partially ordered set $W$ of \emph{worlds}, called the Kripke \emph{frame}, and an indexed family of valuations $(v_w)_{w\in W}$ such that $v_u(A)\leq v_w(A)$ for all $u\leq w$ and propositional variables $A$ (this is called the \emph{persistency condition}).
	The model relation between the $v_u$ and formulas $\varphi$ is defined as before, except in the case of implications, where we set
	\begin{itemize}
		\item $v_u\models\varphi\to \psi$ iff for all $w\geq u$ we have $w\not\models\varphi$ or $w\models\psi$.
	\end{itemize}
	We say that $\varphi$ is satisfied at a world $u$ if $v_u\models\varphi$ and write $u\models\varphi$. If a formula is satisfied at every world then $\mathcal K$ is a \emph{model} for $\varphi$. A formula is valid if every Kripke structure is a model of it. We denote the set of valid formulas with \emph{IPC} (Intuitionistic Propositional Calculus).
\end{definition}
There are many classical tautologies which are not intuitionistically valid, e.g. the law of excluded middle $A\vee\neg A$. On the other hand any intuitionistic theorem is a classical one.

@@

\begin{definition}
	Let $\Sigma$ be a signature. A $\Sigma$-structure $\mathcal{M}$ consists of a non-empty set $M$, the \emph{domain} of $\mathcal{M}$, and an \emph{interpretation} $I$ that assigns
	\begin{itemize}
		\item to each $n$-ary function symbol $f\in\Sigma$ a $n$-ary function $f^I: M^n\to M$.
		\item to each $n$-ary relation symbol $R\in\Sigma$ a $n$-ary relation $R^I\subseteq M^n$.
	\end{itemize}
	A variable assignment $v$ is a function that assigns to each free variable an element $m\in M$. For each free variable $a$ and $m\in M$ we define $$v[m/a](b) = \begin{cases}
	m, &\text{if $b=a$,}\\
	v(b), &\text{otherwise.}
	\end{cases}$$
	Then terms are interpreted as follows:
	\begin{itemize}
		\item $a^{I, v} = v(a)$ for each free variable $a$.
		\item $f(t_1,\dots,t_n)^{I, v} = f^I(t_1^{I, v},\dots, t_n^{I, v})$ for each $n$-ary function symbol $f\in\Sigma$.
	\end{itemize}
	We define a model relation between pairs $\mathcal M, v$ and formulas $\varphi$ as follows
	\begin{itemize}
		\item $\mathcal M, v\not\models\bot$
		\item $\mathcal M, v\models R(t_1,\dots,t_n)$ iff $(t_1^{I, v},\dots,t_n^{I, v})\in R^I$, $\mathcal M, v\models s = t$ iff $s^{I, v} = t^{I, v}$.
		\item $\mathcal M, v\models \varphi\wedge \psi$ iff $\mathcal M, v\models\varphi$ and $\mathcal M, v\models\psi$
		\item $\mathcal M, v\models \varphi\vee\psi$ iff $\mathcal M, v\models\varphi$ or $\mathcal M, v\models\psi$.
		\item $\mathcal M, v\models \varphi\to\psi$ iff $\mathcal M, v\not\models\varphi$ or $\mathcal M, v\models\psi$.
		\item $\mathcal M, v\models\exists x\varphi$ iff there exists $m\in M$ such that $\mathcal M, v[m/a]\models\varphi[a/x]$.
		\item $\mathcal M, v\models\forall x\varphi$ iff for all $m\in M$ we have $\mathcal M, v[m/a]\models\varphi[a/x]$.
	\end{itemize}
	 Here $a$ is a free variable that does not occur in $\varphi$. A $\Sigma$-structure $\mathcal M$ satisfies $\varphi$ if $\mathcal M, v\models\varphi$ for every $v$, we write $\mathcal M\models\varphi$. A formula is valid if every $\Sigma$-structure $\mathcal M$ satisfies it. We denote the set of valid formulas with \emph{CQC} (Classical Quantified Calculus).
\end{definition}

\begin{definition}
	A $\Sigma$-\emph{Kripke structure} $\mathcal{K}$ is a partially ordered set $W$, called the Kripke \emph{frame}, and an indexed family of $\Sigma$-structures $(\mathcal{M}_w)_{w\in W}$ such that for $u\leq w$ we have $M_u\subseteq M_w$, $f^{I_w}|_{M_u} = f^{I_u}$ and $R^{I_w}|_{M_u} = R^{I_u}$.\footnote{Here $f|_M$ denotes the restriction of $f$ to $M$}
	The model relation between $\mathcal M_u$, $v$ (mapping into $M_u$) and formulas $\varphi$ is defined as before, except in the cases of implication and $\forall$-quantification, where we set
	\begin{itemize}
		\item $\mathcal M_u, v\models \varphi\to\psi$ iff for every $w\geq u$ we have $w, v\not\models\varphi$ or $w, v\models\psi$.
		\item $\mathcal M_u, v\models\forall x\varphi$ iff for every $w\geq u$ and $m\in M_w$ we have $w, v[m/a]\models\varphi[a/x]$.
	\end{itemize}
	We write $u, v\models\varphi$ if $\mathcal M_u, v\models \varphi$. $\mathcal{K}$ satisfies $\varphi$ if $u, v\models\varphi$ holds for every world $u$ and variable assignment $v$. $\varphi$ is valid if it is satisfied by every Kripke structure.
	$\varphi$ is valid for some frame (write $W\models\varphi$) if it is satisfied in every Krike structure with that frame. We denote the set of valid formulas with \emph{IQC} (Intuitionistic Quantified Calculus).
\end{definition}
In addition to propositional tautologies there are now quantified formulas which are classically valid but not intuitionistically, e.g., $\neg\forall x A(x)\to \exists x \neg A(x)$.

@@

\begin{definition}
	For formulas $\varphi$ we define the Skolemization $\varphi^S_Z$ and Herbrandization $\varphi^H_Z$ with respect to a set of variables $Z$ by simultaneous induction:
	\begin{itemize}
		\item $A^S_Z = A^H_Z = A$ for each atomic $A$.
		\item $(\varphi\circ\psi)^X_Z = \varphi^X_Z\circ\psi^X_Z$ for $\circ\in\{\wedge, \vee\}$, $X\in\{S, H\}$.
		\item $(\varphi\to\psi)^S_Z = \varphi^H_Z\to \psi^S_Z$ and $(\varphi\to\psi)^H_Z = \varphi^S_Z\to\psi^H_Z$.
		\item $(\forall x\varphi)^S_Z = \forall x(\varphi[a/x]^S_{Z\cup\{a\}}[x/a])$ where $a$ is a new free variable.\\$(\forall x\varphi)^H_Z = \varphi[s(z_1,\dots,z_n)/x]^H_Z$ where $s$ is a new function, $\{z_1\dots z_n\} = Z$.
		\item $(\exists x\varphi)^S_Z = \varphi[s(z_1,\dots,z_n)/x]^S_Z$ where $s$ is a new function, $\{z_1\dots z_n\} = Z$.\\$(\exists x\varphi)^H_Z = \exists x(\varphi[a/x]^H_{Z\cup\{a\}}[x/a])$ where $a$ is a new free variable.
	\end{itemize}
	Let $$\varphi^S = (\exists x_1\dots\exists x_n \varphi[x_1/a_1\dots x_n/a_n])^S_\emptyset$$ and $$\varphi^H = (\forall x_1\dots\forall x_n \varphi[x_1/a_1\dots x_n/a_n])^H_\emptyset$$ where $a_1,\dots,a_n$ are the free variables occurring in $\varphi$.
\end{definition}

@@

Since we could not find a proof of this, one is given in the appendix.

@@

We shall now give an overview of our main results and highlight the key arguments. We start with the propositional case. The important insights are the same as in the first-order case, while it is technically simpler and less cluttered.

@@

The key property that allows us to reduce our encoding to propositional logic, which a priori targets first order logic, is a reduction of counter-models. In particular, given $$\varphi = \left(\bigwedge\mathcal R\wedge\mathcal X\right)\to P$$ as in Lemma~\ref{lemma:propositional-normal-form} we establish:

@@

This allows us to embed intuitionistic propositional logic into classical propositional logic by replacing quantification over worlds by their enumeration.

\begin{lemma}
	For each propositional formula $\varphi$ there is effectively a formula $\varphi^\#$ with $\mathcal |\varphi^\#|\in\mathcal O(|\mathcal R|\cdot\exp(|\mathcal X|\log|\mathcal X|))$ such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid.
\end{lemma}

The exact details can be found in Theorem~\ref{thm:prop-translation}. Instead of expressing correctness at each world of the model, we can express that the worlds along some path in the tree are correct. The branching of paths can be expressed through quantification in a quantified boolean formula (QBF) which gives us the following:

@@

Again, the details can be found in Section~\ref{sec:qbf}. Next we attempt to transport these results to the first order case. However, there are formulas for which only counter-examples with an infinite Kripke frame exist. Therefore, there is no hope of completely eliminating terms representing worlds from the final embedding. This also means that our language necessarily becomes more complex. Still, a number of reductions is possible.
